Demes: a standard format for demographic models

### 2022-05-25 paper.tex

The Demes specification is a formal data model for describing
the properties of populations over time,
along with some metadata and provenance information.
The data model is based on the ubiquitous JSON~\citep{bray2017javascript}
standard, and the structure is formally defined using
JSON Schema~\citep{wright2020json}.
The Demes JSON schema rigorously defines the data model,
specifying the hierarchical structure and the properties that are allowed in different
contexts along with their types and permissible ranges.
This schema, and its accompanying documentation,
fully describe the entities in the model and their
relationships and the required behaviour of implementations.
Since the schema is definitive,
we will not recapitulate the details
here, but instead focus on the high level properties of the model and
the rationale behind key design decisions.

@@

Within a population, we assume that allele frequency dynamics can
be described by the Wright-Fisher model.
Briefly, generations are non-overlapping (all parents
reproduce and die simultaneously), and for allele $i$ currently at frequency $p_i$,
its frequency in the next generation (at birth) is expected to be $p_iw_i/\bar{w}$,
where $w_i$ and $\bar{w}$ are the marginal and
mean fitnesses, respectively, properly weighted according to ancestry proportions.
In this framework,
a forward-time simulation of finite populations is equivalent
to multinomial sampling of allele frequencies each
generation (\citet[][pp 29-31]{burger2000-ul}, \citet[][pp 179-181]{crowkimura1970}),
and a backwards-time (coalescent) simulation follows the approximations
described in \citet{tajima1983evolutionary,hudson1983testing}
and \citet[][chapter 3]{wakeley2008-hd}.
Further, this model assumes ``soft'' selection \citep{christiansen1975hard},
meaning that the dynamics of population sizes
changes are \textit{independent} of the details of individual fitnesses.
As such, this model excludes scenarios such as ``hard selection,''
in which population sizes are dependent on a population's mean fitness, or
stochastic fluctuations in population size, such as interpreting
population sizes as carrying capacities.
Many forwards and backwards time simulators currently implement this model
\citep[e.g.,][]{hudson2002generating,gutenkunst2009inferring,excoffier2011fastsimcoal,kelleher2016efficient,jouganous2017inferring,haller2019slim,thornton2019-nu}.
Each has their own unique method of specifying the details of selection and demography.
The goal of \Demes\ is to provide a common method for the latter task.

@@

Each population has an assigned selfing rate and cloning rate, where each defines
the probability that offspring are generated from one generation to the next by either
self-fertilisation or cloning of an individual.
More specifically, for a given epoch within a population denote the clonal rate by
$\sigma$ and the selfing rate by $S$.
$S$ and $\sigma$ can take any value between zero and one and
can sum to more than one.
Each generation a proportion of offspring
$\sigma$ are expected to be generated through clonal reproduction,
while $1-\sigma$ are expected to arise through sexual reproduction.
Within the sexually-reproduced offspring,
$S$ are born via self-fertilisation while the rest
have parents drawn at random from the previous generation.
Depending on the simulation software, this random drawing of parent may occur
either with or without replacement. When drawing occurs with replacement, a small
amount of ``residual'' selfing is expected, so that the realized selfing probability
is $(1-\sigma)(S + (1-S)/N)$ instead of $(1-\sigma)S$ (so that even with $\sigma=0$
and $S=0$, selfing may still occur with probability $1/N$), although
this effect is negligible in large populations \citep{nordborg1997coalescent}
Simulation software that allow variable rates of selfing are expected to clearly
document the expected behaviour.

@@

Finally, individuals in a population may have parents from a different
population through migrations. These can be defined as continuous migration
rates over time intervals for which populations coexist or through
instantaneous (or pulse) migration events at a given time. Continuous migration
rates are defined as the probability that parents in the ``destination''
population are chosen from the ``source'' population.
On the other hand, pulse
migration events specify the instantaneous replacement of a given fraction of
individuals in a destination population by individuals with parents from a
source population.

This model is, by design, very limited. We do not specify any further details of
individuals or of the processes that happen \emph{within} populations,
because such details are necessarily complicated, vary greatly by
application, and any attempt to encompass such a broad range of biological
processes is doomed to failure. An ``individual'' (i.e., a discrete
organism) is an abstraction that should have a well defined interpretation
in most cases.
This narrow focus on individuals and their groupings into populations
should ensure that the \Demes\ standard is stable, and not subject to
continual development as more and more elaborations of basic processes
are added. We could never hope to fully encompass all the details
of every possible simulation (including ploidy,
genome sizes, recombination, mutation and gene conversion rates,
selection coefficients, and so on) and so we clearly demarcate the
role of the specification by including \emph{none} of them.
Such details can be specified by existing mechanisms used by
simulators. If they vary according to the demography (for example,
we have different recombination maps across populations), then
we can unambiguously define these by referring to the population
identifiers and times defined by the Demes model. Rather than attempting
to embed this information \emph{in} the demographic model, we
can define it externally and refer \emph{to} the demography.

\section{High-level model specification}\label{sec:appendix-spec}

The high-level description of Demes models is focused on the efficiency of
human understanding and the avoidance of errors. We have adopted the widely
used YAML format~\citep{ben2009yaml} as the recommended means of interchanging
Demes models (e.g., Figures~\ref{fig:IM}~and~\ref{fig:tennessen}). YAML is a data
serialisation language with an emphasis on simplicity and which interoperates well
with JSON (indeed, YAML 1.2 is a superset of JSON). We chose YAML over JSON
because although JSON is an excellent format for data interchange, it is
ill-suited for human understanding or manipulation. We also considered other
declarative data exchange formats such as TOML,
but chose YAML because of equivalence with JSON,
its popularity and good software support.
Since the Demes data model is defined in JSON Schema,
however, there is no formal dependency on YAML and implementations may choose
to use JSON directly if they wish (e.g., for greater efficiency).

Structurally, the high-level Demes specification encourages human understanding
by avoiding redundancy in the description where possible and by providing a
mechanism for specifying default values that are inherited hierarchically.
Figure~\ref{fig:IM} shows an example Demes model expressed in both the
high and low-level forms (both in terms of YAML syntax). For values that repeat
across fields, defaults may be used to implicitly assign default values to
fields of the given type.  A default is superseded by an explicitly provided
value if given. Other implicit values are inherited naturally following the
progression of time. For example, if an epoch \texttt{start\_size} is not
provided, it is assumed to be equal to the \texttt{end\_size} of the previous
epoch. This also means that the first epoch of each population must specify the
initial size.

@@

The implicit value inference and default value propagation required to
fully resolve the high-level Demes model description is straightforward
to describe, but still requires significant effort to implement in
software. Moreover, there are many constraints on the data model
(for example, epoch \texttt{end\_time} values should be strictly decreasing
within a deme), which must be enforced.
It would not be reasonable to require every program that
takes Demes models as input to implement this logic, as the programming
effort would be significant (limiting adoption)
and it is likely that if there were many implementations some would differ
in their details (harming the software ecosystem).
% GG: this is currently the case for programs accepting ms command lines. E.g.
%  - MaCS interprets -es proportion as 1-proportion compared to ms.
%  - SCRM uses different options for -ema command.
We therefore define
a standard software entity as part of the specification (the parser),
which performs this task and can be shared by programs that
support Demes as an input. By having relatively few high-quality Demes
parsers available as libraries (ideally, one per programming language),
the probability of divergences from the standard is greatly reduced.

The Demes specification precisely defines the required behaviour of parsers,
which translate the high-level specifications into unambiguous low-level
representations where values have been assigned to all parameters and data
constraints have been checked. The output is formally defined as JSON, but in
practice parsers output an object model that is suitable for the particular
programming language and that can be used directly by the implementing program.
We provide a reference implementation written in Python to resolve any
potential ambiguities and to provide a helpful template for other
implementations, as well as an extensive test suite of examples and the
expected outputs. In addition, we have high-quality parser implementations in
the Python and C languages (published under liberal open source licences),
providing a solid foundation for the software ecosystem.

@@

The Demes specification is designed to describe demographic
models defined by a fixed set of model parameters.
As described in the main text, it does not include
information about estimated confidence intervals or the joint distribution
of parameter values.
While we have opted to restrict Demes to static model descriptions,
allowing parameterized distributions or parameter ranges
is a direction of possible future work.

@@

Demes is designed to interface with a large number of simulation and inference
methods, so we have restricted the specification to only describe
demographic features at the population level. As the underlying discrete
demographic structure is the common thread connecting many widely used
population genetic methods, many features of underlying genome biology are
omitted. Most notably, mutation and recombination rates are left unspecified
(though they can be encoded as metadata in the data structure), as are
sequence length, annotations and other genomic features, and individual ploidy.
Selection and dominance models are absent, as discussed in
Sec.~\ref{sec:appendix-pop-gen-model}.
We also omit information about the sampling of individuals, as we view
the demographic history as independent of the sampling of individuals.
It is therefore left up to simulation and inference tools how to define
genomic parameters and sampling strategies, as each software will have its
own set of constraints and relevant input parameters.

In conclusion,
Demes version 1.0 is intended to provide a stable and minimal specification
that is compatible with a wide range of simulators. To ensure consistency
across implementations, the specification is versioned and we commit to
following semantic versioning guidelines.

@@

\begin{figure}[h!]
    \begin{minipage}{0.44\textwidth}
        \begin{tcolorbox}
            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
        \end{tcolorbox}
        \begin{tcolorbox}
            \includegraphics[width=\textwidth]{fig/IM}
        \end{tcolorbox}
    \end{minipage}\hfill
    \begin{minipage}{0.54\textwidth}
        \begin{tcolorbox}
            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
        \end{tcolorbox}
    \end{minipage}\\
    \caption{
        \label{fig:IM}
        Example isolation-with-migration Demes model. (A) The compact non-redundant
        representation expressed using YAML. (B) The same model in the fully-qualified
        form. (C) A visual representation of the model using \texttt{demesdraw}.
        The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
        but are much more human-readable.
        \aprcomment{Need figure panel labels.}
    }
\end{figure}

>>>>

The Demes specification is a formal data model for describing
the properties of populations over time,
along with some metadata and provenance information.
The data model is based on the ubiquitous JSON~\citep{bray2017javascript}
standard, and formally defined using
JSON Schema~\citep{wright2020json}.
Along with the schema, full technical details of the
of the model are provided in the
online specification document
(\url{https://popsim-consortium.github.io/demes-spec-docs/}).
This specification rigorously defines the data model,
fully describing the entities and their
relationships, and the required behaviour of implementations.
Since the online specification is definitive,
we will not recapitulate the details
here, but instead focus on the high level properties of the model and
the rationale behind key design decisions.

@@

Within a population, we assume that allele frequency dynamics can
be described by the Wright-Fisher model.
Briefly, generations are non-overlapping (all parents
reproduce and die simultaneously), and for allele $i$ currently at frequency $p_i$,
its frequency in the next generation (at birth) is expected to be $p_iw_i/\bar{w}$,
where $w_i$ and $\bar{w}$ are the marginal and
mean fitnesses, respectively, properly weighted according to ancestry proportions.
In this framework,
a forward-time simulation of finite populations is equivalent
to multinomial sampling of allele frequencies each
generation (\citet[][pp 29-31]{burger2000-ul}, \citet[][pp 179-181]{crowkimura1970}),
and a backwards-time (coalescent) simulation follows the approximations
described in \citet{tajima1983evolutionary,hudson1983testing}
and \citet[][chapter 3]{wakeley2008-hd}.
Further, this model assumes ``soft'' selection \citep{christiansen1975hard},
meaning that the dynamics of population sizes
changes are independent of the details of individual fitnesses.
As such, this model excludes scenarios such as ``hard selection,''
in which population sizes are dependent on a population's mean fitness, or
stochastic fluctuations in population size, such as interpreting
population sizes as carrying capacities.
Many forwards and backwards time simulators currently implement this model
\citep[e.g.,][]{hudson2002generating,gutenkunst2009inferring,excoffier2011fastsimcoal,kelleher2016efficient,jouganous2017inferring,haller2019slim,thornton2019-nu}.
Each has their own unique method of specifying the details of selection and demography.
The goal of Demes is to provide a common method for the latter task.

@@

Each population has an assigned selfing rate and cloning rate, where each defines
the probability that offspring are generated from one generation to the next by either
self-fertilisation or cloning of an individual.
More specifically, for a given epoch within a population denote the clonal rate by
$\sigma$ and the selfing rate by $S$.
$S$ and $\sigma$ can take any value between zero and one and
can sum to more than one.
Each generation a proportion of offspring
$\sigma$ are expected to be generated through clonal reproduction,
while $1-\sigma$ are expected to arise through sexual reproduction.
Within the sexually-reproduced offspring,
$S$ are born via self-fertilisation while the rest
have parents drawn at random from the previous generation.
Depending on the simulator, this random drawing of parent may occur
either with or without replacement. When drawing occurs with replacement, a small
amount of ``residual'' selfing is expected, so that the realized selfing probability
is $(1-\sigma)(S + (1-S)/N)$ instead of $(1-\sigma)S$ (so that even with $\sigma=0$
and $S=0$, selfing may still occur with probability $1/N$), although
this effect is negligible in large populations \citep{nordborg1997coalescent}.
Simulators that allow variable rates of selfing are expected to clearly
document the expected behaviour.

@@

Individuals in a population may have parents from a different
population through migrations. These can be defined as continuous migration
rates over time intervals for which populations coexist or through
instantaneous (or pulse) migration events at a given time. Continuous migration
rates are defined as the probability that parents in the ``destination''
population are chosen from the ``source'' population.
On the other hand, pulse
migration events specify the instantaneous replacement of a given fraction of
individuals in a destination population by individuals with parents from a
source population.

\section{Human Data Model}\label{sec:appendix-spec}

The Demes Human Data Model (HDM) is focused on the efficiency of
human understanding and the avoidance of errors. We have adopted the widely
used YAML format~\citep{ben2009yaml} as the recommended means of interchanging
Demes models (e.g., Figures~\ref{fig:IM}~and~\ref{fig:tennessen}). YAML is a data
serialisation language with an emphasis on simplicity and which interoperates well
with JSON (indeed, YAML 1.2 is a superset of JSON). We chose YAML over JSON
because although JSON is an excellent format for data interchange, it is
ill-suited for human understanding or manipulation. We also considered other
declarative data exchange formats such as TOML,
but chose YAML because of equivalence with JSON,
its popularity and good software support.
Since the Demes data model is defined in JSON Schema,
however, there is no formal dependency on YAML and implementations may choose
to use JSON directly if they wish (e.g., for greater efficiency).

Structurally, the HDM encourages human understanding
by avoiding redundancy in the description where possible and by providing a
mechanism for specifying default values that are inherited hierarchically.
Figure~\ref{fig:IM} shows an example Demes model expressed in both the
HDM and Machine Data Model (MDM) forms (both in terms of YAML syntax).
For values that repeat
across fields, defaults may be used to implicitly assign default values to
fields of the given type.  A default is superseded by an explicitly provided
value if given. Other implicit values are inherited naturally following the
progression of time. For example, if an epoch \texttt{start\_size} is not
provided, it is assumed to be equal to the \texttt{end\_size} of the previous
epoch. This also means that the first epoch of each population must specify the
initial size.

@@

The implicit value inference and default value propagation required to
fully resolve a Demes HDM description into the corresponding
MDM form is straightforward
to describe, but still requires significant effort to implement in
software. Moreover, there are many constraints on the data model
(for example, epoch \texttt{end\_time} values should be strictly decreasing
within a deme), which must be enforced.
It would not be reasonable to require every program that
takes Demes models as input to implement this logic, as the programming
effort would be significant (limiting adoption)
and it is likely that if there were many implementations some would differ
in their details (harming the software ecosystem).
% GG: this is currently the case for programs accepting ms command lines. E.g.
%  - MaCS interprets -es proportion as 1-proportion compared to ms.
%  - SCRM uses different options for -ema command.
We therefore define
a standard software entity as part of the specification (the parser),
which performs this task and can be shared by programs that
support Demes as an input. By having relatively few high-quality Demes
parsers available as libraries (ideally, one per programming language),
the probability of divergences from the standard is greatly reduced.

The Demes specification precisely defines the required behaviour of parsers,
which translate a model described in the HDM into the MDM, where values have been
assigned to all parameters and data
constraints have been checked. The output is formally defined as JSON, but in
practice parsers output an object model that is suitable for the particular
programming language and that can be used directly by the implementing program.
We provide a reference implementation written in Python to resolve any
potential ambiguities and to provide a helpful template for other
implementations, as well as an extensive test suite of examples and the
expected outputs. In addition, we have high-quality parser implementations in
the Python and C languages (published under liberal open source licences),
providing a solid foundation for the software ecosystem.

@@

The Demes specification is designed to describe demographic
models defined by a fixed set of model parameters.
As described in the main text, it does not include
information about estimated confidence intervals or the joint distribution
of parameter values. In this section we describe the rationale
for this design decision.

@@

Demes is designed to interface with a large number of simulation and inference
methods, so we have restricted the specification to only describe
demographic features at the population level.
Features of the underlying genome biology are omitted,
including mutation and recombination rates,
genome annotations, ploidy, sequence annotations, and so on.
Selection and dominance models are absent, as discussed in
Sec.~\ref{sec:appendix-pop-gen-model}, and we do not
specify how individuals are sampled.

The model is, by design, very limited. We do not specify details of
individuals or of the processes that happen within populations,
because such details are necessarily complicated, vary greatly by
application, and attempts to encompass such a broad range of biological
processes are unlikely to succeed.
The narrow focus on individuals and their groupings into populations
should ensure that the Demes standard is stable, and not subject to
continual development as more and more elaborations of basic processes
are added. Demes is intended to provide \emph{part} of a
simulation model specification; other parts of the specification,
such as sampling strategies, or parameters that vary by time
or population can \emph{refer to} this well defined and unambigous
description of the demography.

@@

\begin{figure}[h!]
    \begin{minipage}{0.44\textwidth}
        \begin{tcolorbox}
            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM.yaml}
        \end{tcolorbox}
        \begin{tcolorbox}
            \includegraphics[width=\textwidth]{fig/IM}
        \end{tcolorbox}
    \end{minipage}\hfill
    \begin{minipage}{0.54\textwidth}
        \begin{tcolorbox}
            \inputminted[fontsize=\scriptsize,linenos,numbersep=5pt]{yaml}{models/IM-resolved.yaml}
        \end{tcolorbox}
    \end{minipage}\\
    \caption{
        \label{fig:IM}
        Example isolation-with-migration Demes model. (A) The Human Data Model
        representation expressed using YAML. (B) The same model in the
        Machine Data Model form.
        (C) A visual representation of the model using \texttt{demesdraw}.
        The YAML descriptions in (A) and (B) correspond exactly to a JSON description,
        but are much more human-readable.
        \aprcomment{Need figure panel labels.}
    }
\end{figure}
